---
title: "NCCS Data Exploration"
author: "Matt Peyton"
date: "9/25/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(scales)
```

## About the National Center for Charitable Statistics (NCCS)


### Getting Started with NCCS Data

[Beginners Guide to NCCS Data](https://nccs.urban.org/sites/default/files/2018-12/Guide%20to%20Using%20NCCS%20Data.pdf)

[NCCS Data: R Users Guide](https://nccs.urban.org/publication/nccs-data-r-users-guide#nccs-data-r-users-guide)

**3 Types of Files** 

* IRS Business Master Files (BMF)  
* NCCS Core Files (Core)  
* IRS Statistics of Income Exempt Organizations Sample Files (SOI)  

### IRS Business Master Files (BMF)

**Updated**

* BMF datasets are generally released multiple times per year; at least one per year is available going back to 1995
* Files are updated by the IRS on a regular basis, and downloaded to the NCCS Data Archive several times per year
* The information in the BMF is almost always the most up to date information about these organizations and their filing statuses.

**Contains**

* Each file represents the number of organizations registered with the IRS during the month/year indicated
* The BMF contains records both for active and inactive organizations, as well as registered organizations operating below the minimum
requirements for filing a Form 990 or 990-EZ. 

**Does Not Contain**

* The number of total fields from the Form 990 are very limite
* Financial fields are limited solely to total revenue and total assets
* Many Parts of the Form 990 are not represented at all 

### NCCS Core Files

[Data Dictionary](https://nccs-data.urban.org/showDD.php?ds=core)

**Updated**

* Released once per year, the NCCS Core Files represent the entire population of active, reporting organizations of the given type filing within a given circa year

**Contains**

* Thus, each Core File contains multiple fiscal years’ worth of information, but only one record for each EIN.
* This data is aggregated to show an accurate snapshot of actual
nonprofit activity within the given year, accounting for the varying filing patterns and fiscal years of different nonprofits.
* The NCCS Core Files contain a far wider selection of financial fields from the Form 990 than the BMF, including detailed information on revenue, expenses, and assets. 
* Further, the Core files are subject to the NCCS validation procedure before release, making the data more accurate than that
contained in the BMF. 

## Get CORE Files

```{r echo=FALSE}

#Code to download and prepare National Center for Charitable Statistics Core (PC) Files
#Brice McKeever
#Center on Nonprofits and Philanthropy
#The Urban Institute
#3/9/18

#For information on the National Center for Charitable Statistics, see: http://nccs.urban.org/

#Load relevant libraries
library(tidyverse)
library(httr)

#The following function will download and prepare NCCS Core Files for analysis.
#The function requires the following fields:
#     year: year of data (as of this writing, available years include 1989-2015 for most data sets)
#     type: type of data, given as follows (include quotes)
#           "pc" = 501(c)(3) public charities
#           "pf" = 501(c)(3) private foundations
#           "co" = all other 501(c) organizations


#Create function to download core files
getcorefile <- function(coreyear, coretype) {
    #ensure coretype is lowercase
    coretype <- tolower(coretype)
    
    #make sure user entered pc, pf, or co
    check <- isTRUE(coretype %in% c("pc", "pf", "co"))
    if (check == "FALSE"){
        stop("Please make sure to enter \"pc\" (for public charities), \"pf\" (for private foundations)', or \"co\" (for other 501(c) organizations)")
    }
    
    #create URL for download based on year of file
    if(coretype =="co"){
        URL <- paste("http://nccs-data.urban.org/data/core/", as.character(coreyear), "/coreco.core", as.character(coreyear),as.character(coretype), ".csv", sep ="") 
    }
    if(coretype  %in% c("pc", "pf")) {
        URL <- paste("http://nccs-data.urban.org/data/core/", as.character(coreyear), "/nccs.core", as.character(coreyear),as.character(coretype), ".csv", sep ="")
    }
    
    #get raw data
    rawcore <- GET(as.character(URL))
    if (rawcore$status_code != 200){
        stop("Please check database: it is possible that Core year/type combination you entered does not exist.  See http://nccs-data.urban.org/data.php?ds=core for list of all available datasets, and enter a valid year/type combination")
    }
    stop_for_status(rawcore)
    
    #read data as CSV file format, with proper formatting
    
    #Note 1: the code below reads in most fields from the relevant Core file.  However, users should consider commenting out lines
    #that they do not require.  By default, fields used in NCCS validation processes are not included in output file.
    #Note 2: The code below assumes column names for most recent years of the Core files.  Older editions of the NCCS Core Files
    #may have other column names.  Please consult the relevant data dictionaries (at http://nccs-data.urban.org/data-dictionaries.php)
    #for more information, and change as neecssary.  
    
    
    #parse PC and CO files
    if(coretype  %in% c("pc", "co")) {
        cfile <- content(rawcore, type = "text/csv", 
                         col_types = cols_only(  EIN = col_character(),
                                                 ACCPER = col_character(),
                                                 ACTIV1 = col_character(),
                                                 ACTIV2 = col_character(),
                                                 ACTIV3 = col_character(),
                                                 ADDRESS = col_character(),
                                                 AFCD = col_character(),
                                                 ASS_BOY = col_double(),
                                                 ASS_EOY = col_double(),
                                                 #BLOCK = col_character(),
                                                 BOND_BOY = col_double(),
                                                 BOND_EOY = col_double(),
                                                 CENSUSTRACT = col_character(),
                                                 CITY = col_character(),
                                                 CLASSCD = col_character(),
                                                 COMPENS = col_double(),
                                                 #COMPENSP = col_double(),
                                                 CONT = col_double(),
                                                 CONTACT = col_character(),
                                                 #CONTP = col_double(),
                                                 DEDUCTCD = col_character(),
                                                 DIREXP = col_double(),
                                                 EOSTATUS = col_character(),
                                                 EPOSTCARD = col_character(),
                                                 EXPS = col_double(),
                                                 #EXPSP = col_double(),
                                                 FILENAME = col_character(),
                                                 FIPS = col_character(),
                                                 FISYR = col_double(),
                                                 #FISYRP = col_double(),
                                                 FNDNCD = col_character(),
                                                 FRCD = col_character(),
                                                 FUNDBAL = col_double(),
                                                 FUNDFEES = col_double(),
                                                 #FUNDFEESP = col_double(),
                                                 FUNDINC = col_double(),
                                                 FUNDINCP = col_double(),
                                                 GFTGRNTSRCVD170 = col_double(),
                                                 GOODS = col_double(),
                                                 GRPROF = col_double(),
                                                 #GRPROFP = col_double(),
                                                 GRREC = col_double(),
                                                 GRRECMEM = col_double(),
                                                 GRRECOTH = col_double(),
                                                 GRRECPUB = col_double(),
                                                 GRSINC170 = col_double(),
                                                 GRSINCFNDRSNG = col_double(),
                                                 GRSINCGAMING = col_double(),
                                                 GRSRCPTSADMISSN509 = col_double(),
                                                 GRSRCPTSRELATED170 = col_double(),
                                                 GRSRNTSPRSNL = col_double(),
                                                 GRSRNTSREAL = col_double(),
                                                 INITFEES = col_double(),
                                                 #INPRIOR = col_integer(),
                                                 #INPRIORSRC = col_character(),
                                                 INVENTG = col_double(),
                                                 INVINC = col_double(),
                                                 #INVINCP = col_double(),
                                                 LATITUDE = col_double(),
                                                 LESSDIRFNDRSNG = col_double(),
                                                 LESSDIRGAMING = col_double(),
                                                 LEVEL1 = col_character(),
                                                 LEVEL2 = col_character(),
                                                 LEVEL3 = col_character(),
                                                 LEVEL4 = col_character(),
                                                 LIAB_BOY = col_double(),
                                                 LIAB_EOY = col_double(),
                                                 LONGITUDE = col_double(),
                                                 MAJGRPB = col_character(),
                                                 #MANUALLY_FIXED = col_character(),
                                                 MRTG_BOY = col_double(),
                                                 MRTG_EOY = col_double(),
                                                 MSA_NECH = col_character(),
                                                 NAICS = col_character(),
                                                 NAME = col_character(),
                                                 #NCCSKEY = col_character(),
                                                 #NCCSKEY2 = col_character(),
                                                 NETA_BOY = col_double(),
                                                 NETGNLS = col_double(),
                                                 NETINC = col_double(),
                                                 NETINCFNDRSNG = col_double(),
                                                 NETINCGAMING = col_double(),
                                                 #NETINCP = col_double(),
                                                 NETRENT = col_double(),
                                                 #NETRENTP = col_double(),
                                                 NTEE1 = col_character(),
                                                 NTEECC = col_character(),
                                                 NTEECONF = col_character(),
                                                 NTEEFINAL = col_character(),
                                                 NTEEFINAL1 = col_character(),
                                                 NTEEIRS = col_character(),
                                                 NTEESRC = col_character(),
                                                 NTMAJ10 = col_character(),
                                                 NTMAJ12 = col_character(),
                                                 NTMAJ5 = col_character(),
                                                 ORGCD = col_character(),
                                                 OTHCHGS = col_double(),
                                                 OTHINC = col_double(),
                                                 OTHINCP = col_double(),
                                                 OTHSAL = col_double(),
                                                 #OTHSALP = col_double(),
                                                 OUTNCCS = col_character(),
                                                 OutNCCS = col_character(),
                                                 OUTREAS = col_character(),
                                                 PAYTAX = col_double(),
                                                 PAYTAXP = col_double(),
                                                 PMSA = col_character(),
                                                 PROGREV = col_double(),
                                                 #PROGREVP = col_double(),
                                                 Q78A = col_character(),
                                                 RANDNUM = col_double(),
                                                 REASON = col_character(),
                                                 RENTEXP = col_double(),
                                                 RENTINC = col_double(),
                                                 RETEARN = col_double(),
                                                 RETE_BOY = col_double(),
                                                 RNTLEXPNSPRSNL = col_double(),
                                                 RNTLEXPNSREAL = col_double(),
                                                 RNTLINCPRSNL = col_double(),
                                                 RNTLINCREAL = col_double(),
                                                 ROYALTSINC = col_double(),
                                                 RULEDATE = col_character(),
                                                 SALEOTHE = col_double(),
                                                 SALEOTHG = col_double(),
                                                 SALEOTHN = col_double(),
                                                 #SALEOTHNP = col_double(),
                                                 SALESECN = col_double(),
                                                 #SALESECNP = col_double(),
                                                 SALESEXP = col_double(),
                                                 SECUR = col_double(),
                                                 SEC_NAME = col_character(),
                                                 SOIYR = col_double(),
                                                 SOURCE = col_character(),
                                                 SPEVTG = col_double(),
                                                 SRVCSVAL170 = col_double(),
                                                 SRVCSVAL509 = col_double(),
                                                 STATE = col_character(),
                                                 STYEAR = col_double(),
                                                 SUBCD = col_character(),
                                                 SUBSECCD = col_character(),
                                                 SUBTOTSUPPINC509 = col_double(),
                                                 TAXPER = col_integer(),
                                                 #TAXPERP = col_double(),
                                                 TOTGFTGRNTRCVD509 = col_double(),
                                                 TOTREV = col_double(),
                                                 TOTREV2 = col_double(),
                                                 #TOTREVP = col_double(),
                                                 TOTSUPP509 = col_double(),
                                                 TXEXMPTBNDSPROCEEDS = col_double(),
                                                 TXREVNUELEVIED170 = col_double(),
                                                 TXREVNUELEVIED509 = col_double(),
                                                 UNSECUREDNOTESEND = col_double(),
                                                 #VALIDATION_STATE = col_double(),
                                                 ZIP = col_character(),
                                                 ZIP5 = col_character()))
    }
    
    #parse PF file
    if(coretype == "pf"){
        cfile <- content(rawcore, type = "text/csv", 
                         col_types = cols_only(
                             EIN = col_character(),
                             ACCPER = col_character(),
                             ACTIV1 = col_character(),
                             ACTIV2 = col_character(),
                             ACTIV3 = col_character(),
                             ADDRESS = col_character(),
                             AFCD = col_character(),
                             #ASS_CODE = col_character(),
                             BLOCK = col_character(),
                             CENSUSTRACT = col_character(),
                             CITY = col_character(),
                             CLASSCD = col_character(),
                             CONTACT = col_character(),
                             #DEDUCTCD = col_character(),
                             #EOSTATUS = col_integer(),
                             FIPS = col_character(),
                             FISYR = col_integer(),
                             FNDNCD = col_character(),
                             FRCD = col_character(),
                             #INC_CODE = col_character(),
                             LATITUDE = col_double(),
                             LEVEL1 = col_character(),
                             LEVEL2 = col_character(),
                             LEVEL3 = col_character(),
                             LEVEL4 = col_character(),
                             LONGITUDE = col_double(),
                             MAJGRPB = col_character(),
                             #MANUALLY_FIXED = col_character(),
                             MSA_NECH = col_character(),
                             NAICS = col_character(),
                             NAME = col_character(),
                             #NCCSKEY = col_double(),
                             #NCCSKEY2 = col_double(),
                             NTEE1 = col_character(),
                             NTEECC = col_character(),
                             NTEECONF = col_character(),
                             NTEEFINAL = col_character(),
                             NTEEFINAL1 = col_character(),
                             NTEEIRS = col_character(),
                             NTEESRC = col_character(),
                             NTMAJ10 = col_character(),
                             NTMAJ12 = col_character(),
                             NTMAJ5 = col_character(),
                             ORGCD = col_character(),
                             OUTNCCS = col_character(),
                             OutNCCS = col_character(),
                             OUTREAS = col_character(),
                             P10MIR = col_double(),
                             P10TASST = col_double(),
                             P11DISTR = col_double(),
                             P13UNDST = col_double(),
                             P14A4942 = col_double(),
                             P14ASVLA = col_double(),
                             P14ASVLB = col_double(),
                             P14ASVLC = col_double(),
                             P14ASVLD = col_double(),
                             P14B4942 = col_double(),
                             P14C4942 = col_double(),
                             P14D4942 = col_double(),
                             P14ENDWA = col_double(),
                             P14ENDWB = col_double(),
                             P14ENDWC = col_double(),
                             P14ENDWD = col_double(),
                             P14GINVA = col_double(),
                             P14GINVB = col_double(),
                             P14GINVC = col_double(),
                             P14GINVD = col_double(),
                             P14NADJA = col_double(),
                             P14NADJB = col_double(),
                             P14NADJC = col_double(),
                             P14NADJD = col_double(),
                             P14PSUPA = col_double(),
                             P14PSUPB = col_double(),
                             P14PSUPC = col_double(),
                             P14PSUPD = col_double(),
                             P14QDISA = col_double(),
                             P14QDISB = col_double(),
                             P14QDISC = col_double(),
                             P14QDISD = col_double(),
                             P14T4942 = col_double(),
                             P14TASVL = col_double(),
                             P14TENDW = col_double(),
                             P14TGINV = col_double(),
                             P14TNADJ = col_double(),
                             P14TPSUP = col_double(),
                             P14TQDIS = col_double(),
                             P14TSUPA = col_double(),
                             P14TSUPB = col_double(),
                             P14TSUPC = col_double(),
                             P14TSUPD = col_double(),
                             P14TTSUP = col_double(),
                             P1ADMEXP = col_double(),
                             P1CHAEXP = col_double(),
                             P1CONTPD = col_double(),
                             P1DIVID = col_double(),
                             P1EXCREV = col_double(),
                             P1GINVPF = col_double(),
                             P1GOODS = col_double(),
                             P1GRENTS = col_double(),
                             P1INTREV = col_double(),
                             P1INVEXP = col_double(),
                             P1INVRV = col_double(),
                             P1NADINC = col_double(),
                             P1NADJRV = col_double(),
                             P1NETINV = col_double(),
                             P1NGASTS = col_double(),
                             P1OFCOMP = col_double(),
                             P1OTHINC = col_double(),
                             P1TADJEX = col_double(),
                             P1TCONT = col_double(),
                             P1TEXMEX = col_double(),
                             P1TINVEX = col_double(),
                             P1TOTEXP = col_double(),
                             P1TOTREV = col_double(),
                             P2CASHBV = col_double(),
                             P2CRPBND = col_double(),
                             P2CRPSTK = col_double(),
                             P2EYASST = col_double(),
                             P2EYOTIN = col_double(),
                             P2GVTINV = col_double(),
                             P2MORTG = col_double(),
                             P2MTGNTS = col_double(),
                             P2TASFMV = col_double(),
                             P2TINVSC = col_double(),
                             P2TLIABL = col_double(),
                             P2TOTAST = col_double(),
                             P3EYTFND = col_double(),
                             P6DOMORG = col_character(),
                             P6ESTTX = col_double(),
                             P6EXMPF = col_character(),
                             P6TEXCTX = col_double(),
                             P6TX4940 = col_double(),
                             P6TX511 = col_double(),
                             P6TXCRDT = col_double(),
                             P6TXINV = col_double(),
                             P6TXPNLT = col_double(),
                             P6TXRFD = col_double(),
                             P6TXSUBA = col_double(),
                             P6TXWERR = col_double(),
                             P6TXWTH = col_double(),
                             P7BUSINT = col_character(),
                             P7DISBOR = col_character(),
                             P7DISPAY = col_character(),
                             P7DISSAL = col_character(),
                             P7DISSER = col_character(),
                             P7DISTRA = col_character(),
                             P7ELECT = col_character(),
                             P7INDGRT = col_character(),
                             P7JEOINV = col_character(),
                             P7LIQUID = col_character(),
                             P7ORGGRT = col_character(),
                             P7OTHPUR = col_character(),
                             P7PAYGVT = col_character(),
                             P7POFCLM = col_character(),
                             P7POLIT = col_character(),
                             P7PROPAG = col_character(),
                             P7UNDINC = col_character(),
                             PMSA = col_character(),
                             RANDNUM = col_double(),
                             #RECCODE = col_character(),
                             RULEDATE = col_character(),
                             SEC_NAME = col_character(),
                             SOURCE = col_character(),
                             STATE = col_character(),
                             STYEAR = col_character(),
                             SUBSECCD = col_character(),
                             TAXPER = col_character(),
                             #TFLD = col_integer(),
                             #VALIDATION_STATE = col_double(),
                             ZIP = col_character(),
                             ZIP5 = col_character()
                         ))
    }
    
    
    
    #convert all variable names to uppercase
    names(cfile) <- toupper(names(cfile))
    
    #write output to local drive
    #Note: Users should STRONGLY consider saving the data locally to avoid repeated downloads of the data
    write.csv(cfile, file = as.character(paste("core", as.character(coreyear), as.character(coretype), ".csv", sep="")))
    
    #return output to R for immediate exploration
    return(cfile)
}

#Examples using function defined above:

#Download 2015 Core PC File:
#core2015pc <- getcorefile(2015, "pc")

#Download 2015 Core CO File:
#core2015co <- getcorefile(2015, "co")

#Download 2014 Core PF file:
#core2014pf <- getcorefile(2014, "pf")


#Examples of flawed code:

#Will not run, as "zz" is not a valid Core file type
#coretest <- getcorefile(2015, "ZZ")

#Will not run, as 2045 is not a valid year for PC data
#coretest2 <- getcorefile(2045, "pc")

```

```{r}
core2015pc <- getcorefile(2015, "pc")
```

```{r}

core2015pc %>%
    ggplot(aes(x=ASS_EOY, y=EXPS, color=ACCPER)) +
    geom_point(alpha=.2)+
    scale_x_log10() +
    scale_y_log10()
    #scale_x_continuous(labels = comma, limits=c(0,20000000000)) +
    #scale_y_continuous(labels = comma, limits=c(0,20000000000)) 

```